Another DSL primer

### 2013-12-25 docs/fsml/paper.tex
Enabled negative test cases and added force semantics to megamodel execution. Some code clean-up needed soon, but this is a pretty useful state of affairs now.

0.00002 as of 25 November 2013.

@@

@@

See \autoref{languages/fsml/ok.pro} for the Prolog specification
listing the named constraints that we consider here. Each constraint
gives rise to a separate predicate considered below. Here is short
summary. Constraint \emph{fsmSingleInitial} is meant to ensure that
there is only a single state declaration for an initial
state. Constraint \emph{fsmDistinctIds} is meant to ensure that the
state ids of the state declarations are distinct. Constraint
\emph{fsmResolvable} is meant to ensure that all referenced target
states are declared. Constraint \emph{fsmDeterministic} is meant
to ensure that each given input uniquely defines the target state to
be transitioned to, if any. Constraint \emph{fsmReachable} is meant to
ensure that all declared states are reachable from the
initial state. In \autoref{languages/fsml/ok.pro}, all constraint
predicates are surround by an application of the meta-predicate
\emph{require/1}, which is simply there for better error
reporting. That is, if the argument of \emph{require/1} fails, then
this is reported immediately, as opposed to simply propagating failure
silently.

@@

@@

@@

@@

@@

@@

All constraints are specified in
\autoref{languages/fsml/ok/initial.pro}--\autoref{languages/fsml/ok/closure.pro}. The
constraints are relatively simple to express using Prolog's
meta-predicate \emph{findall/3} which serves here for queries into the
terms of the abstract syntax. For instance, in
\autoref{languages/fsml/ok/initial.pro}, we find all state ids from
state declarations with `true' for the initial component. Once this
set is retrieved that the resulting list is of length 1; thus, there
is exactly one initial.

@@

@@

\codefigure{%
Generic interface for action handers}{%
languages/fsml/java/Handler.java}{%
java}

\codefigure{%
A trivial interpretation of the turnstile actions}{%
languages/fsml/java/TurnstileHandler.java}{%
java}

\autoref{languages/fsml/java/Handler.java} shows a generic, i.e.,
FSM-independent interface for defining handlers for FSM
actions. \autoref{languages/fsml/java/TurnstileHandler.java} shows a
trivial implementation of said interface for the turnstile
example. The actions simply print out the names of the actions, but it
is clear that arbitrary functionality can be plugged into FSM
execution in this manner.

\codefigure{%
Generic class for FSM execution}{%
languages/fsml/java/Stepper.java}{%
java}

\autoref{languages/fsml/java/Stepper.java} shows essentially the
runtime for FSMs. This is generic code which is appropriately
parameterized in the types for states, inputs, and actions. The class
relies on a table for holding the transitions in the form of a map
from states to maps from inputs to pairs of optional actions and
states. There is an appropriate \emph{add} method which makes it easy
to insert into this non-basic container. The class is also prepared to
keep track of and make use of a handler for the involved actions. The
class is called \emph{Stepper} because its key method is the one for
performing a step (a transition) based on an actual input item.

\codefigure{%
Configuration of a stepper for the turnstile example}{%
languages/fsml/java/TurnstileStepper.java}{%
java}

\autoref{languages/fsml/java/TurnstileStepper.java} shows the
generated code for the turnstile-specific \emph{Stepper} subclass. The
class instantiates the generic parameters of the \emph{Stepper} class
to the turnstile-specific enum types for states, inputs, and actions,
which were shown earlier. Also, the class encodes all transitions of
the turnstile FSM via appropriate calls to the \emph{add} method, as
part of the constructor. The initial state is also set up and the
turnstile-specific handler is communicated as well.

\codefigure{%
Code generation for stepper subclass}{%
languages/fsml/to-java/stepper.pro}{%
prolog}

\autoref{languages/fsml/to-java/stepper.pro} shows the code generation
predicate for stepper subclasses. The predicate refers to more
elements of the Java language: several expression forms (e.g.,
`null'), assignment statements, method calls, and class
declarations. In a final step, the constructed abstract syntactical
representation is passed to the pretty printer for Java.

@@

>>>>

0.00003 as of 25 December 2013.

@@

Various technical details are separated out into the appendix.

@@

See \autoref{languages/fsml/ok.pro} for the Prolog specification
listing the named constraints that we consider here. Each constraint
gives rise to a separate predicate considered below. Here is short
summary. Constraint \emph{fsmSingleInitial} is meant to ensure that
there is exactly one initial state declaration. Constraint
\emph{fsmDistinctIds} is meant to ensure that the state ids of the
state declarations are distinct. Constraint \emph{fsmResolvable} is
meant to ensure that all referenced target states are
declared. Constraint \emph{fsmDeterministic} is meant to ensure that
each given input uniquely defines the target state to be transitioned
to, if any. Constraint \emph{fsmReachable} is meant to ensure that all
declared states are reachable from the initial state. In
\autoref{languages/fsml/ok.pro}, all constraint predicates are
surround by an application of the meta-predicate \emph{require/1},
which is simply there for better error reporting. That is, if the
argument of \emph{require/1} fails, then this is reported immediately,
as opposed to simply propagating failure silently.

@@

\codefigure{%
A negative test case: \autoref{languages/fsml/sample.fsml} with a second initial state}{%
languages/fsml/tests/initialNotOk.fsml}{%
fsml}

@@

\codefigure{%
A negative test case: \autoref{languages/fsml/sample.fsml} with a
double declaration of state `locked'}{%
languages/fsml/tests/idsNotOk.fsml}{%
fsml}

@@

\codefigure{%
A negative test case: \autoref{languages/fsml/sample.fsml} with a
unresolvable state reference `unlockkked'}{%
languages/fsml/tests/resolutionNotOk.fsml}{%
fsml}

@@

\codefigure{%
A negative test case: \autoref{languages/fsml/sample.fsml} with a
nondeterministic transition; see the double occurrence of `ticket' in
state `locked'}{%
languages/fsml/tests/determinismNotOk.fsml}{%
fsml}

@@

\codefigure{%
A negative test case: \autoref{languages/fsml/sample.fsml} with an
unreachable state `unlocked'}{%
languages/fsml/tests/reachabilityNotOk.fsml}{%
fsml}

@@

All constraints are specified in
\autoref{languages/fsml/ok/initial.pro}--\autoref{languages/fsml/ok/closure.pro}.
We also show negative test cases that hence illustrate violation of
the constraints; these cases are derived by minimalistic mutation of
the sample FSM of \autoref{languages/fsml/sample.fsml}. The
constraints are relatively simple to express using Prolog's
meta-predicate \emph{findall/3} which serves here for queries into the
terms of the abstract syntax. For instance, in
\autoref{languages/fsml/ok/initial.pro}, we find all state ids from
state declarations with `true' for the initial component. Once this
set is retrieved that the resulting list is of length 1; thus, there
is exactly one initial.

@@

\codefigure{%
A possible input sequence for FSM of
\autoref{languages/fsml/sample.fsml}}{%
languages/fsml/sample.input}{%
prolog}

\codefigure{%
The output for 
\autoref{languages/fsml/sample.input}}{%
languages/fsml/sample.output}{%
prolog}

Let us set up a test case for the FSM of
\autoref{languages/fsml/sample.fsml}. That is, a valid input sequence
for the FSM is shown in \autoref{languages/fsml/sample.input} and the
corresponding output in terms of actions and state changes is shown in
\autoref{languages/fsml/sample.input}. We shall define a semantics
that indeed transforms the input into the output.

@@

\codefigure{%
Generic interface for action handers}{%
languages/fsml/java/HandlerBase.java}{%
java}

\codefigure{%
A trivial interpretation of the turnstile actions}{%
languages/fsml/java/Handler.java}{%
java}

\autoref{languages/fsml/java/HandlerBase.java} shows a generic, i.e.,
FSM-independent interface for defining handlers for FSM
actions. \autoref{languages/fsml/java/Handler.java} shows a
trivial implementation of said interface for the turnstile
example. The actions simply print out the names of the actions, but it
is clear that arbitrary functionality can be plugged into FSM
execution in this manner.

\codefigure{%
Generic class for FSM execution}{%
languages/fsml/java/StepperBase.java}{%
java}

\autoref{languages/fsml/java/StepperBase.java} shows essentially the
runtime for FSMs. This is generic code which is appropriately
parameterized in the types for states, inputs, and actions. The class
relies on a table for holding the transitions in the form of a map
from states to maps from inputs to pairs of optional actions and
states. There is an appropriate \emph{add} method which makes it easy
to insert into this non-basic container. The class is also prepared to
keep track of and make use of a handler for the involved actions. The
class is called \emph{StepperBase} because its key method is the one for
performing a step (a transition) based on an actual input item.

\codefigure{%
Configuration of a stepper for the turnstile example}{%
languages/fsml/java/Stepper.java}{%
java}

\autoref{languages/fsml/java/Stepper.java} shows the generated code
for the turnstile-specific subclass of the \emph{StepperBase}
class. The generic parameters of \emph{StepperBase} are instantiated
to the turnstile-specific enum types for states, inputs, and actions,
which were shown earlier. Also, the class encodes all transitions of
the turnstile FSM via appropriate calls to the \emph{add} method, as
part of the constructor. The initial state is also set up and the
turnstile-specific handler is communicated as well.

\codefigure{%
Code generation for stepper class}{%
languages/fsml/to-java/stepper.pro}{%
prolog}

\autoref{languages/fsml/to-java/stepper.pro} shows the code generation
predicate for steppers. The predicate refers to more elements of the
Java language: several expression forms (e.g., `null'), assignment
statements, method calls, and class declarations. In a final step, the
constructed abstract syntactical representation is passed to the
pretty printer for Java.

@@

\codefigure{%
A negative test case: \autoref{languages/fsml/sample.fsml} with a syntax error; see `innnitial'}{%
languages/fsml/tests/parserError.fsml}{%
fsml}

\codefigure{%
A negative test case: an input for
\autoref{languages/fsml/sample.fsml} which contains an invalid input symbol}{%
languages/fsml/tests/illegalSymbol.input}{%
prolog}

\codefigure{%
A negative test case: an input for
\autoref{languages/fsml/sample.fsml} which contains an infeasible
input symbol in the pool position---the initial state `locked' does
not accept the `mute' input}{%
languages/fsml/tests/infeasibleSymbol.input}{%
prolog}

\section{Additional test cases for the specification}
\label{A:testing}

Let us add a few more negative test cases in addition to the
illustrations of constraint violation of \S\ref{S:ok}.  A simple
parsing error is caused by the input of
\autoref{languages/fsml/tests/parserError.fsml}.
\autoref{languages/fsml/tests/illegalSymbol.input} and
\autoref{languages/fsml/tests/infeasibleSymbol.input} demonstrate two
major scenarios for failure of `simulation' according to
\S\ref{S:semantics}.
